Efficient Computation of Extended Surface Sources

### 2023-01-26 repro/papers/GEM22/submitR2/response.tex

\documentclass[12pt]{article}

\begin{document}
\title{Response to Comments on ``Efficient Computation of Extended Surface Sources'', Revision 1, submitted to IJGE}
\maketitle

Once again, I thank the reviewers for a very helpful comments and suggestions, and for amazingly diligent typo detection. Comments and respponses follow.

\noindent {\bf Editor:}

@@

Dear Author, 
Thanks for this revised version of your  study. Both reviewers acknowledge the significant improvement of this new version. They still point some typos, and comments/questions I think would be beneficial to consider in a further revised version of your study before we accept it for publication. See these comments appended below. 
Thanks a lot for your work and this interesting study, 
Best regards,
Ludovic Métivier.

Response: Once again the reviews have been quite helpful. I have tried to address each point raised by the reviewers.

\noindent {\bf Reviewer 1:}

Comment: The manuscript has been considerably revised. The revised version is much better. The early introduction of the examples and the overview of the time reversal method significantly improve the readability. I would like to recommend the manuscript for publication. 
I have, however, two questions and I have detected typos (I am afraid that there are many more).

Response: I have tried to answer the questions, and correct the typos without however introducing as many new ones!

@@

Comment 1. what are the blue and red rays in Figure 3 ?
the blue rays start from the source, the red rays start from a point located just below the source. As a result, some red and blue rays are very close to each other.

Response: I have created a clearer ray diagram overlain on the bulk modulus color scale plot (Figure 3) and added another figure and some discussion to explain the ray geometry. The red and blue rays start at the same source point and are simply different parts of the same point source ray field. The diving wave example is similar to that depicted in Figure 4.20a in Sheriff and Geldart, {\em Exploration Seismology}. It features a homogeneous upper layer, a middle layer in which the velocity increases rapidly, and a homogeneous lower layer, with smooth transitions between the layers. As Sheriff and Geldart say, ``A velocity gradient in a layer that is substantially higher than the velocity gradient above the bed, as in fig. 4.20a, causes a very sharp increase in ray path curvature and a folding back of the time-distance curve. Such a triplication of branches of the time-distance curve (fig. 4.20c) is usually difficult to see because the later branches become lost in the later cycles of the earlier arrivals.'' To show clearly how this triplication happens, I've included a plot of arrival times in my introduction of this example (new Figure 4b). The part of the arrival time curve in red corresponds to the those red rays in the ray plot (Figure 3) that refract back to the receiver surface, and similarly for the blue rays. Note that the red branch is close enough to the direct arrival time that its chief effect is an apparent wavelet change in the direct arrival. Both red and blue branches are really diving wave arrival times, but only the waveform arrival corresponding to the blue branch can be isolated by a simple mute.

Comment: on page 23, I do not understand the following sentence:
The function $a_0$ is a frequency-independent envelope, equal to 1 where h is non-zero.

Response: This section skipped over a number of steps, which I have included in the revision. I hope that this more detailed account explains this idea.

\noindent {\bf Reviewer 2:}

Comment: Overall the paper is much improved and I now found it relatively easy to follow and understand.  I particularly liked the 'steps' introduced as a way to summarize what you've done so far and to give the reader further help in actually implementing these ideas.  (I also liked the roadmap section at the beginning.)  I also found the new figures much easier to compare and understand.
There are a few areas where I think things could still be somewhat clearer, which are:

Response: Thank you for your comments, which inspired these changes in R1.

@@

Comment 1. In the introduction p 3 line 44: Perhaps say in a few words why you isolate the diving rays.  Also later when you show your ray picture (Figure 3) with the diving waves, first I think I see red rays beside the blue ones, and second just out of curiosity, what is your definition of diving wave, because I see things that turn back to the surface that you do not deem diving rays.

Response: You are of course correct that the ``things that turn back'' are another diving wave branch. I've considerably revised the figure, its caption, and the discussion in the text, intending to answer all of these questions.

@@

Comment 2. You relate what you are doing to photo-acoustic tomography, which is neat and certainly was the motivation.  As I read the paper though, I can't help thinking that what you are doing is a lot like running RTM back to the source.  I expect that other readers may also wonder what is really fundamentally different between RTM and this approach, so I think that adding something to make this comparison explicit would be helpful.

Response: this is a very good question. I have added a paragraph describing the role of the D-to-N map in ``true amplitude RTM'', aka approximate Born inversion. In fact it's quite similar to the role described in this paper for the inverse surface source problem. In both cases the core idea is to backpropagate the actual data field. In the present case, that is a transmitted field; in true-amplitude migration, it's the scatterered field. Reconstructing these from the data is the key to inversion in both cases, and the D-to-N map is the key to reconstruction.

Comment 3. In figure 2 and the lens example in general, why is the source you used not on the source surface?  Perhaps I have misunderstood the meaning of the source surface, and that could be clarified.  I guess that the source should be below that surface so the rays come into it from below and then leave it 'from above' and enter the receiver surface from below?   I am sure that one could phrase that much better than I have here, but as I was reading it was odd to see the source below what's called the source surface.

Response: I could have placed the source on the source surface. I created a parallel sequence of experiments for the lens example using that configuration. However a point source cannot be reconstructed, at least not with the simple transmission geometry used in this paper, since a lot of its energy propagates out of aperture and is not available in the inversion. You can reconstruct a source that generates the data, but it isn't a point source - if the velocity is right, it is more or less a dip-filtered version of the point source. Instead, I chose to somewhat artificially create a source that can actually be reconstructed fairly accurately in this example: the rays carrying significant energy actually make it to the receiver surface within the time limit.

The diving wave example uses a point source on the source surface. You will notice that the reconstructed source (Figure 9b) appears to be a dip-filtered point source, as expected.

@@

Comment 4, In Figure 3, there are some rays that look weird to me, at around (8000,1000) and (12000,1000) This is just a gradient so I am expecting more-or-less circular wavefronts.  I don't know that this really requires a comment or anything, but maybe a mention as to what is going on there.

Response: Figure 3 and surrounding discussion needed some tuning up. Note that the gradient-to-constant transition is smoothed, and the gradient is large enough to generate a triplication. Having homogeneous layers top and bottom really messes things up in an interesting way. Sheriff and Geldart, {\em Exploration Geophysics}, discussion of fig. 4.20a, mention this behaviour. There are segments of circles in there, but overall these rays are not close to circles. I think the description now in the text, and the inclusion of the arrival time plot, will make things clear.

@@

Comment 5. p 11 line 32, what do you mean by 'must be localized' in which variables?

Response: I have added a bit of discussion that I hope will justify and explain this assertion. The PAT construction does not apply to the surface source inversion problem, since there is no bounding surface on all of which the pressure field is measured/known. I get around this by localizing in space-time. For short-time propagation, the signal does not propagate far from the source surface, so an artificial bounding surface can be added on which the field is automatically zero. So the field can be viewed as the solution of a local pressure boundary value problem, for a little while. Then it can be converted to a local solution of a problem with surface source, as described. Having taken that step, you can simply get rid of the artificial boundary, since the field vanishes near it. If the pressure field on the source (or receiver) surface is only non-zero for a short time, that's it - just let it propagate in forward time. Finally, any pressure distribution on the source surface can be broken up in to a sum of terms, each only non-zero for a short interval. Apply the short-time-window construction to each summand, and add up the resulting fields.

@@

Comment 6. Figure 6b has quite a different source than that shown in Figure 2b.  I think that when you talk about Figure 6 it would be good to re-iterate why you are getting a different source here (I think velocity vs pressure, but remembering that many people only look at the pictures this will be confusing).

Response: I think this is actually explained in the caption of Figure 7 (formerly Figure 6), also in the text discussion. This is the result of inversion {\em in the wrong velocity}. The data (Figure 2a or 2c) is inverted here in a homogeneous model - so the source does not look at all like that (2b) used to generate the data in the lens model.

\noindent {\bf Typos and other corrections:}

I am {\em very} grateful to both referees for taking the trouble to list many of the embarassingly extensive typos. I believe that I have fixed all of these, and some others which we won't mention. I only hope I have not created many new typos...

I also corrected a number if errors in figure and equation references, some of which may once have been right but no longer were.

The section ``Computing $\Lambda$'' needed revision, as I had glossed over the fact that the approximate inverse is only a {\em right} approximate inverse - necessarily, since the modeling operator $S$, as I have defined it, can have small singular values corresponding to out-of-aperture transmission, or possibly even an outright null space. A consequence is that the approximation of $\Lambda$ computed in this section is accurate up to an error that is neglible after application of $S$, and that needed to be stated clearly.

\end{document}

### 2023-01-26 repro/papers/GEM22/submitR2/paper.tex
maybe final

Source extension is one approach to avoiding this``cycle-skipping''
obstacle. It consists in imposing the wave equation as a soft
constraint, allowing the source field $\bff$ to have more degrees of
freedom than is permitted by a faithful model model of the seismic
experiment, and constraining these additional degrees of freedom by
means of an additive quadratic penalty modifying the problem
\ref{eqn:ols}:
\begin{equation}
\label{eqn:esi}
\mbox{choose } {\bf c}, \bff \mbox{ to minimize } \|PL[{\bf c}]^{-1}\bff -d \|^2 + \alpha^2 \|A\bff\|^2 
\end{equation}
The linear operator $A$ penalizes deviation from known (or assumed)
characteristics of the source function - its null space consists of
feasible (or ``physical'') source models.

@@

As noted earlier (and visible in Figure \ref{fig:olensrays0}), the acoustic field in this example satisfies the ray
conditions on which the time reversal construction is based: The rays
are incoming at $\Sigma_s$, outgoing at $\Sigma_r$ and all rays
carrying significant energy at points of $\Sigma_r$ pass over
$\Sigma_s$. The outward unit normal at $\Sigma_s$ is ${\bf n}_s ={\bf
  e}_z=(0,1)$; at $\Sigma_r$, it's ${\bf n}_r =-{\bf
  e}_z=(0,-1)$. So the source on the receiver surface $\Sigma_r$, input to reverse time propagation
(system \ref{eqn:awerecglob}, generating the fields $(\hat{p},\hat{\bv})$) is
$h_r=2P_rv_z$ (the field $P_rv_z$ is depicted in Figure \ref{fig:dfwdvzlh0}). After reverse time propagation , the approximate
inversion $\hat{h}_s=2P_s\hat{v}_z$ is shown in Figure
\ref{fig:dinvhslh0}. It differs considerably from the source field used to
generate the data (Figure \ref{fig:dsrcphh0}), however when used as the
source in forward modeling (system \ref{eqn:awesrcglob}) {\em with the
  same material parameters used in the inversion}, it generates a
pressure gather on $\Sigma_r$ (Figure \ref{fig:drerecplh0}) very
closely 
approximating the input gather (Figure \ref{fig:dfwdplh0}). The
difference is plotted on the same color scale in Figure
\ref{fig:ddiffrecplh0}.

\begin{figure}
  \centering
  \subfloat[\label{fig:dfwdvzlh0}]{\includegraphics[width=0.45\textwidth]{dfwdvzlh0.pdf}}
  \subfloat[\label{fig:dinvhslh0}]{\includegraphics[width=0.45\textwidth]{dinvhslh0.pdf}}
  \caption{(a) Normal velocity data corresponding to pressure data in
    Figure \ref{fig:dfwdplh0}. (b) Source function resulting from time-reversal inversion
    of pressure data in Figure \ref{fig:dfwdplh0}. Propagation in homogeneous
    bulk modulus model (different from the model used to generate the
    data!). Plotted with same color scale as Figure \ref{fig:dsrcphh0}.}
\end{figure}

@@

Observe that in principle it's not necessary to
supply the normal velocity at the source (or receiver) surface, since
it's determined locally by the pressure on the surface, or causal (or
anti-causal) fields. This follows from the observation that the system
\ref{eqn:aweloc} has a unique solution, a by-product of the
conservation of energy (see for example \cite{Lax:PDENotes}, section
4.3). That is, in the language of that discussion, knowledge of $p$ on
$\Sigma$ for $0\le t \le \tau$ determines $(\tilde{p},\tilde{\bv})$ in
the entire domain $\Omega^+$, $0\le t \le \tau$. Therefore knowledge
of $p$ on $\Sigma$ for $0\le t \le \tau$ determines
$\bv \cdot {\bf n}$, up to a negligible error. As in the last section,
it follows that knowledge of $p$ on $\Sigma$, any $t$, determines a
source $h$ on $\Sigma$ for which the pressure field $\tilde{p}$
(solution of the system \ref{eqn:awesrcmod}) differs negligibly from
$p$.

@@

\subsection{Time Reversal as Adjoint State}
The middle step is solution of the reverse-time system
\ref{eqn:awerecglob} for the field $(\hat{p},\hat{\bv})$, with input
pressure source
$h_r$ on $\Sigma_r$ and output $P_s\hat{p}$ on $\Sigma_s$. To
understand the significance of this mapping, it is just as easy, and
will be useful later, to include a velocity source $f_r$, that is,
modify the second equation in \ref{eqn:awerecglob} by adding $f_r {\bf
  n}_r \delta_{\Sigma_r}$, so that it becomes a reverse-time version
of the basic system \ref{eqn:awepm}. Recall the definition of energy
in the acoustic field $(p,\bv)$:
\begin{equation}
  \label{eqn:energydef}
   E[p,\bv](t) = \frac{1}{2} \int \,dx\,dy\,dz\, \left(\frac{p(\bx,t)^2}{\kappa(\bx)} + \rho(\bx)\bv(\bx,t)
   \cdot \bv(\bx,t)\right)
\end{equation}
This is a quadratic form. Polarization makes a bilinear form, into
which one can plug a pair of fields. Use the solutions $(p,\bv)$ of
the forward-time system \ref{eqn:awepm} for one of the pair, and
the solution $(\hat{p},\hat{\bv})$ of system \ref{eqn:awerecglob}
(augmented with the velocity surface source) for the other. Since $(p,\bv)$ vanishes
for large negative time, $(\hat{p},\hat{\bv})$ for large positive
time, 
\[
0 = 
\left(\int\, dx\,dy\,dz\, \frac{p \hat{p}}{\kappa} +  
\rho \bv \cdot \hat{\bv} \right)|_{t \rightarrow \infty}
-
\left(\int\, dx\,dy\,dz\, \frac{p \hat{p}}{\kappa} +  \rho \bv \cdot \hat{\bv} \right)|_{t \rightarrow -\infty}
\]
\[
= 
\int_{-\infty}^{\infty} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p \hat{p}}{\kappa} +  \rho \bv \cdot \hat{\bv} \right)
\]
Carry out the differentiation under the integral sign, replace time derivatives by space derivatives and
source terms using the equations of motion in the systems
\ref{eqn:awepm} and \ref{eqn:awerecglob}, and integrate by parts to
move all of the space derivatives onto $p$ and $\bv$.  Most of the
remaining expressions cancel, leaving
\[
  = \int_{-\infty}^{\infty}\,dt \, \left[
    \int\,d\sigma_s(\bx)\,(h_sP_s\hat{p}+f_sP_s(\hat{\bv}\cdot{\bf
      n}_s)) +
    \int\,d\sigma_r(\bx)\,(h_rP_rp+f_rP_r(\bv\cdot{\bf n}_r)) \right]
\]
Here $\sigma_s$ and $\sigma_r$ are the area elements on $\Sigma_s$ and
$\Sigma_r$. These integrals are inner products of time traces on $\Sigma_s$ and $\Sigma_r$. Denote these inner products by $\langle \cdot,\cdot
\rangle_s$ and $\langle \cdot,\cdot
\rangle_r$ respectively. Then the above is written in terms of the
vector modeling operator ${\cal S}$ (definition \ref{eqn:fwd}) as 
\[
  = \langle (h_s,f_s)^T, (P_s\hat{p},P_s(\hat{\bv}\cdot{\bf
    n}_s))^T \rangle_s+ \langle (h_r,f_r)^T, {\cal S}(h_s,f_s)^T \rangle_r,
\]
whence identify the adjoint of ${\cal S}$ as
\begin{equation}
  \label{eqn:vecadj}
  {\cal S}^T(h_r,f_r) = -(P_s\hat{p},P_s(\hat{\bv}\cdot{\bf
    n}_s))^T.
\end{equation}
Specializing to the pressure modeling operator $S$ (equation
\ref{eqn:sdef}),
\begin{equation}
  \label{eqn:padj}
  S^T = (\Pi_0 {\cal S} \Pi_0^T)^T = \Pi_0 {\cal S}^T \Pi_0^T
\end{equation}
so to calculate the adjoint of $S$ applied to $h_r$, solve system
\ref{eqn:awerecglob} (with $f_r=0$) and extract the pressure at $\Sigma_s$:
\[
  S^T(h_r) = -P_s \hat{p}.
\]

@@

\subsection{Approximate symmetry of $\Lambda$}
Except in special cases, $\Lambda_{s,r}$ is not
symmetric. However, it is {\em approximately symmetric} in the
high-frequency sense. This fact follows from a high-frequency
asymptotic analysis, a sketch of which follows for the special case of
a planar source surface
$\Lambda_s=\{z=z_s\}$, as in the two examples. Since the same reasoning
applies to $\Lambda_r$, I will omit the subscript for the remainder of
this subsection.

@@

The upshot of this calculation is that to leading order in frequency,
the normal velocity corresponding to a pressure wave packet differs by a position- and slowness-dependent
factor. Accounting for the factor of 2 difference between the source
amplitude and velocity trace, this calculation may be summarized:
\begin{equation}
  \label{eqn:lamwp}
  \Lambda (a_0e^{i\omega(t-x\xi-y\eta)}) = \lambda
  (a_0e^{i\omega(t-x\xi-y\eta)}),
\end{equation}
where $\lambda(x,y,t,\xi,\eta,\omega)$ is
defined by the asymptotic series in $\omega$ developed above. Its dominant term,
$O(1)$ (order 0) in $\omega$, is
\[
  \lambda^0(x,y,t,\xi,\eta,\omega) = 2\frac{( c^{-2}(x,y,z)-
    \xi^2+\eta^2)^{1/2}}{\rho} + O(1/\omega)
\]
\begin{equation}
  \label{eqn:lamsymb}
= 2(\kappa
\rho)^{-1/2}\left(1-\frac{\kappa(k_x^2+k_y^2)}{\rho \omega^2}\right)^{1/2} + O(1/\omega).
\end{equation}
In the final expression, the slownesses $\xi,\eta$ have been replaced
by the Fourier variables $k_x=x\xi, k_y=y\eta$. Note that as a
function of $k_x,k_y,\omega$, $\lambda^0$ is homogeneous of order
0. It is easy to see that the $N$th term of the asymptotic series
defining $\lambda$ is of homogeneous order $-N$.

@@

Direct computation of the pressure-to-source operator $\Lambda_s$, for instance, by solving
\ref{eqn:awepm} and reading off $P_sv_{z}$, turns out to be
numerically ill-behaved. An alternate approach begins with the
observation that a pressure gather on $\Sigma_r$ produced by a
pressure source $h_s$ on $\Sigma_s$ can also be well-approximated by
the field from a velocity source. To see this, recall the first major
step in the time reversal construction, that a solution of the wave
equation near $\Sigma_s$ can be represented as the solution of a
boundary value problem with $p$ given on $\Sigma_s$ (system
\ref{eqn:aweloc}). The derivation combined the unique determination of
the field by its pressure boundary values on an enclosing domain with
the consequences of the standing assumptions about rays carrying high
frequency energy. Precisely the same logic could have been applied to
the normal velocity boundary values, as these also uniquely determine
the field in an enclosing domain. Proceeding further, a source
representation based on continuous extension of normal velocity, with
a jump in the pressure follows by logic precisely analogous to that
leading to the system \ref{eqn:awesrc}: instead, obtain a field
$(\tilde{p},\tilde{\bv})$ differing negligibly from the field of the
same name in the earlier discussion, solving 
\begin{eqnarray}
\label{eqn:awesrcvzmod}
  \frac{1}{\kappa}\frac{\partial\tilde{p}}{\partial t} & = & 
                                                      - \nabla \cdot \tilde{\bv}, \nonumber \\
  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & f_s {\bf n}\delta_{\Sigma}- \nabla \tilde{p}, \nonumber \\
  \tilde{p} & = & 0 \mbox{ for }  t \ll 0, \nonumber \\
  \tilde{\bv} & = & {\bf 0} \mbox{ for }  t \ll  0.       
\end{eqnarray}
(that is, system \ref{eqn:awepm} with $h=0$) with $f_s = 2P_sp$.

@@

\begin{figure}
  \centering
  \subfloat[\label{fig:dsrcphh0}]{\includegraphics[width=0.45\textwidth]{dsrcphh0.pdf}}
  \subfloat[\label{fig:dsrcvzhh0}]{\includegraphics[width=0.45\textwidth]{dsrcvzhh0.pdf}}
  \caption{(a) Pressure gather for Lens example on source surface
    $\Sigma_s=\{z=1000\mbox{ m}\}$, generated by point source at
    $(3500,3500)$ m. (b) Normal velocity gather for Lens example on source surface
    $\Sigma_s=\{z=1000\mbox{ m}\}$, generated by point source at
    $(3500,3500)$ m.}
\end{figure}

\begin{figure}
  \centering
  \subfloat[\label{fig:dfwdplh0bis}]{\includegraphics[width=0.32\textwidth]{dfwdplh0.pdf}}
  \subfloat[\label{fig:daltplh0}]{\includegraphics[width=0.32\textwidth]{daltplh0.pdf}} 
  \subfloat[\label{fig:ddifffwdaltplh0}]{\includegraphics[width=0.32\textwidth]{ddifffwdaltplh0.pdf}}
  \caption{(a) Pressure gather on receiver surface
    $\Sigma_r=\{z=1000\mbox{ m}\}$ due to extended pressure source = 2
    $\times$ normal velocity gather shown in Figure
    \ref{fig:dsrcvzhh0}. This is the same gather as shown in Figure
    \ref{fig:dfwdplh0}, and the extended pressure source has been
    shown in Figure \ref{fig:dhshh0}.
    (b) Pressure gather on receiver surface
    $\Sigma_r=\{z=1000\mbox{ m}\}$ due to extended velocity source = 2
    $\times$ pressure gather shown in Figure
    \ref{fig:dsrcvzhh0}. (c) Difference (b)-(a), plotted on same color
    scale.}
\end{figure}

>>>>

Source extension is one approach to avoiding this``cycle-skipping''
obstacle. It consists in imposing the wave equation as a soft
constraint, allowing the source field $\bff$ to have more degrees of
freedom than is permitted by a faithful model of the seismic
experiment, and constraining these additional degrees of freedom by
means of an additive quadratic penalty modifying the problem
\ref{eqn:ols}:
\begin{equation}
\label{eqn:esi}
\mbox{choose } {\bf c}, \bff \mbox{ to minimize } \|PL[{\bf c}]^{-1}\bff -d \|^2 + \alpha^2 \|A\bff\|^2 
\end{equation}
The linear operator $A$ penalizes deviation from known (or assumed)
characteristics of the source function - its null space consists of
feasible (or ``physical'') source models.

@@

As noted earlier (and visible in Figure \ref{fig:olensrays0}), the acoustic field in this example satisfies the ray
conditions on which the time reversal construction is based: The rays
are incoming at $\Sigma_s$, outgoing at $\Sigma_r$ and all rays
carrying significant energy at points of $\Sigma_r$ pass over
$\Sigma_s$. The outward unit normal at $\Sigma_s$ is ${\bf n}_s ={\bf
  e}_z=(0,1)$; at $\Sigma_r$, it's ${\bf n}_r =-{\bf
  e}_z=(0,-1)$. So the source on the receiver surface $\Sigma_r$, input to reverse time propagation
(system \ref{eqn:awerecglob}, generating the fields $(\hat{p},\hat{\bv})$) is
$h_r=2P_rv_z$ (the field $P_rv_z$ is depicted in Figure \ref{fig:dfwdvzlh0}). After reverse time propagation , the approximate
inversion $\hat{h}_s=2P_s\hat{v}_z$ is shown in Figure
\ref{fig:dinvhslh0}. It differs considerably from the source field used to
generate the data (Figure \ref{fig:dhshh0}), however when used as the
source in forward modeling (system \ref{eqn:awesrcglob}) {\em with the
  same material parameters used in the inversion}, it generates a
pressure gather on $\Sigma_r$ (Figure \ref{fig:drerecplh0}) very
closely 
approximating the input gather (Figure \ref{fig:dfwdplh0}). The
difference is plotted on the same color scale in Figure
\ref{fig:ddiffrecplh0}.

\begin{figure}
  \centering
  \subfloat[\label{fig:dfwdvzlh0}]{\includegraphics[width=0.45\textwidth]{dfwdvzlh0.pdf}}
  \subfloat[\label{fig:dinvhslh0}]{\includegraphics[width=0.45\textwidth]{dinvhslh0.pdf}}
  \caption{(a) Normal velocity data corresponding to pressure data in
    Figure \ref{fig:dfwdplh0}. (b) Source function resulting from time-reversal inversion
    of pressure data in Figure \ref{fig:dfwdplh0}. Propagation in homogeneous
    bulk modulus model (different from the model used to generate the
    data!). Plotted with same color scale as Figure \ref{fig:dhshh0}.}
\end{figure}

@@

Observe that in principle it's not necessary to
supply the normal velocity at the source (or receiver) surface, since
it's determined locally by the pressure on the surface, for causal (or
anti-causal) fields. This follows from the observation that the system
\ref{eqn:aweloc} has a unique solution, a by-product of the
conservation of energy (see for example \cite{Lax:PDENotes}, section
4.3). That is, in the language of that discussion, knowledge of $p$ on
$\Sigma$ for $0\le t \le \tau$ determines $(\tilde{p},\tilde{\bv})$ in
the entire domain $\Omega^+$, $0\le t \le \tau$. Therefore knowledge
of $p$ on $\Sigma$ for $0\le t \le \tau$ determines
$\bv \cdot {\bf n}$, up to a negligible error. As in the last section,
it follows that knowledge of $p$ on $\Sigma$, any $t$, determines a
source $h$ on $\Sigma$ for which the pressure field $\tilde{p}$
(solution of the system \ref{eqn:awesrcmod}) differs negligibly from
$p$.

@@

\subsection{Time Reversal as Adjoint State}
The middle step is solution of the reverse-time system
\ref{eqn:awerecglob} for the field $(\hat{p},\hat{\bv})$, with input
pressure source
$h_r$ on $\Sigma_r$ and output $P_s\hat{p}$ on $\Sigma_s$. To
understand the significance of this mapping, it is just as easy, and
will be useful later, to include a velocity source $f_r$, that is,
modify the second equation in \ref{eqn:awerecglob} by adding $f_r {\bf
  n}_r \delta_{\Sigma_r}$, so that it becomes a reverse-time version
of the basic system \ref{eqn:awepm}. Recall the definition of energy
in the acoustic field $(p,\bv)$:
\begin{equation}
  \label{eqn:energydef}
   E[p,\bv](t) = \frac{1}{2} \int \,dx\,dy\,dz\, \left(\frac{p(\bx,t)^2}{\kappa(\bx)} + \rho(\bx)\bv(\bx,t)
   \cdot \bv(\bx,t)\right)
\end{equation}
This is a quadratic form. Polarization makes a bilinear form, into
which one can plug a pair of fields. Use the solutions $(p,\bv)$ of
the forward-time system \ref{eqn:awepm} for one of the pair, and
the solution $(\hat{p},\hat{\bv})$ of system \ref{eqn:awerecglob}
(augmented with the velocity surface source) for the other. Since $(p,\bv)$ vanishes
for large negative time, $(\hat{p},\hat{\bv})$ for large positive
time, 
\[
0 = 
\left(\int\, dx\,dy\,dz\, \frac{p \hat{p}}{\kappa} +  
\rho \bv \cdot \hat{\bv} \right)|_{t \rightarrow \infty}
-
\left(\int\, dx\,dy\,dz\, \frac{p \hat{p}}{\kappa} +  \rho \bv \cdot \hat{\bv} \right)|_{t \rightarrow -\infty}
\]
\[
= 
\int_{-\infty}^{\infty} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p \hat{p}}{\kappa} +  \rho \bv \cdot \hat{\bv} \right)
\]
Carry out the differentiation under the integral sign, replace time derivatives by space derivatives and
source terms using the equations of motion in the systems
\ref{eqn:awepm} and \ref{eqn:awerecglob}, and integrate by parts to
move all of the space derivatives onto $p$ and $\bv$.  Most of the
remaining expressions cancel, leaving
\[
  = \int_{-\infty}^{\infty}\,dt \, \left[
    \int\,d\sigma_s(\bx)\,(h_sP_s\hat{p}+f_sP_s(\hat{\bv}\cdot{\bf
      n}_s)) +
    \int\,d\sigma_r(\bx)\,(h_rP_rp+f_rP_r(\bv\cdot{\bf n}_r)) \right]
\]
Here $\sigma_s$ and $\sigma_r$ are the area elements on $\Sigma_s$ and
$\Sigma_r$. These integrals are inner products of time traces on $\Sigma_s$ and $\Sigma_r$. Denote these inner products by $\langle \cdot,\cdot
\rangle_s$ and $\langle \cdot,\cdot
\rangle_r$ respectively. Then the above is written in terms of the
vector modeling operator ${\cal S}$ (definition \ref{eqn:fwd}) as 
\[
  = \langle (h_s,f_s)^T, (P_s\hat{p},P_s(\hat{\bv}\cdot{\bf
    n}_s))^T \rangle_s+ \langle (h_r,f_r)^T, {\cal S}(h_s,f_s)^T \rangle_r,
\]
whence identify the adjoint of ${\cal S}$ as
\begin{equation}
  \label{eqn:vecadj}
  {\cal S}^T(h_r,f_r) = -(P_s\hat{p},P_s(\hat{\bv}\cdot{\bf
    n}_s))^T.
\end{equation}
Specializing to the pressure modeling operator $S$ (equation
\ref{eqn:sdef}),
\begin{equation}
  \label{eqn:padj}
  S^T = (\Pi_0 {\cal S} \Pi_0^T)^T = \Pi_0 {\cal S}^T \Pi_0^T
\end{equation}
so to calculate the adjoint of $S$ applied to $h_r$, solve system
\ref{eqn:awerecglob} (with $f_r=0$) and extract the pressure at $\Sigma_s$:
\[
  S^Th_r = -P_s \hat{p}.
\]

@@

\subsection{Approximate symmetry of $\Lambda$}
Except in special cases, $\Lambda_{s,r}$ is not
symmetric. However, it is {\em approximately symmetric} in the
high-frequency sense. This fact follows from a high-frequency
asymptotic analysis, a sketch of which follows for the special case of
a planar source surface
$\Sigma_s=\{z=z_s\}$, as in the two examples. Since the same reasoning
applies to $\Lambda_r$, I will omit the subscript for the remainder of
this subsection.

@@

The upshot of this calculation is that to leading order in frequency,
the normal velocity corresponding to a pressure wave packet differs by a position- and slowness-dependent
factor. Accounting for the factor of 2 difference between the source
amplitude and velocity trace, this calculation may be summarized:
\begin{equation}
  \label{eqn:lamwp}
  \Lambda (a_0e^{i\omega(t-x\xi-y\eta)}) = \lambda
  (a_0e^{i\omega(t-x\xi-y\eta)}),
\end{equation}
where $\lambda(x,y,t,\xi,\eta,\omega)$ is
defined by the asymptotic series in $\omega$ developed above. Its dominant term,
$O(1)$ (order 0) in $\omega$, is
\[
  \lambda^0(x,y,t,\xi,\eta,\omega) = 2\frac{( c^{-2}(x,y,z)-
    (\xi^2+\eta^2))^{1/2}}{\rho} + O(1/\omega)
\]
\begin{equation}
  \label{eqn:lamsymb}
= 2(\kappa
\rho)^{-1/2}\left(1-\frac{\kappa(k_x^2+k_y^2)}{\rho \omega^2}\right)^{1/2} + O(1/\omega).
\end{equation}
In the final expression, the slownesses $\xi,\eta$ have been replaced
by the Fourier variables $k_x=\omega\xi, k_y=\omega\eta$. Note that as a
function of $k_x,k_y,\omega$, $\lambda^0$ is homogeneous of order
0. It is easy to see that the $N$th term of the asymptotic series
defining $\lambda$ is of homogeneous order $-N$.

@@

Direct computation of the pressure-to-source operator $\Lambda_s$, for instance, by solving
\ref{eqn:awepm} and reading off $P_sv_{z}$, turns out to be
numerically ill-behaved. An alternate approach begins with the
observation that a pressure gather on $\Sigma_r$ produced by a
pressure source $h_s$ on $\Sigma_s$ can also be well-approximated by
the field from a velocity source. To see this, recall the first major
step in the time reversal construction, that a solution of the wave
equation near $\Sigma_s$ can be represented as the solution of a
boundary value problem with $p$ given on $\Sigma_s$ (system
\ref{eqn:aweloc}). The derivation combined the unique determination of
the field by its pressure boundary values on an enclosing domain with
the consequences of the standing assumptions about rays carrying high
frequency energy. Precisely the same logic could have been applied to
the normal velocity boundary values, as these also uniquely determine
the field in an enclosing domain. Proceeding further, a source
representation based on continuous extension of normal velocity, with
a jump in the pressure follows by logic precisely analogous to that
leading to the system \ref{eqn:awesrc}: instead, obtain a field
$(\tilde{p},\tilde{\bv})$ differing negligibly from the field of the
same name in the earlier discussion, solving 
\begin{eqnarray}
\label{eqn:awesrcvzmod}
  \frac{1}{\kappa}\frac{\partial\tilde{p}}{\partial t} & = & 
                                                      - \nabla \cdot \tilde{\bv}, \nonumber \\
  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & f_s {\bf n}\delta_{\Sigma_s}- \nabla \tilde{p}, \nonumber \\
  \tilde{p} & = & 0 \mbox{ for }  t \ll 0, \nonumber \\
  \tilde{\bv} & = & {\bf 0} \mbox{ for }  t \ll  0.       
\end{eqnarray}
(that is, system \ref{eqn:awepm} with $h=0$) with $f_s = 2P_sp$.

@@

\begin{figure}
  \centering
  \subfloat[\label{fig:dsrcphh0}]{\includegraphics[width=0.45\textwidth]{dsrcphh0.pdf}}
  \subfloat[\label{fig:dsrcvzhh0}]{\includegraphics[width=0.45\textwidth]{dsrcvzhh0.pdf}}
  \caption{(a) Pressure gather for Lens example on source surface
    $\Sigma_s=\{z=3000\mbox{ m}\}$, generated by point source at
    $(3500,3500)$ m. (b) Normal velocity gather for Lens example on source surface
    $\Sigma_s=\{z=3000\mbox{ m}\}$, generated by point source at
    $(3500,3500)$ m.}
\end{figure}

\begin{figure}
  \centering
  \subfloat[\label{fig:dfwdplh0bis}]{\includegraphics[width=0.32\textwidth]{dfwdplh0.pdf}}
  \subfloat[\label{fig:daltplh0}]{\includegraphics[width=0.32\textwidth]{daltplh0.pdf}} 
  \subfloat[\label{fig:ddifffwdaltplh0}]{\includegraphics[width=0.32\textwidth]{ddifffwdaltplh0.pdf}}
  \caption{(a) Pressure gather on receiver surface
    $\Sigma_r=\{z=1000\mbox{ m}\}$ due to extended pressure source = 2
    $\times$ normal velocity gather shown in Figure
    \ref{fig:dsrcvzhh0}. This is the same gather as shown in Figure
    \ref{fig:dfwdplh0}, and the extended pressure source has been
    shown in Figure \ref{fig:dhshh0}.
    (b) Pressure gather on receiver surface
    $\Sigma_r=\{z=1000\mbox{ m}\}$ due to extended velocity source = 2
    $\times$ pressure gather shown in Figure
    \ref{fig:dsrcphh0}. (c) Difference (b)-(a), plotted on same color
    scale.}
\end{figure}
